Criticisms of the IHME COVID-19 model

### 2020-04-25 text/20/ihme-covid-model.Rmd
First complete draft of ihme-covid-model

@@

To avoid too much confusion, I will use the term "IHME model" to refer to the COVID-19 prediction model developed by the Institute for Health Metrics and Evaluation and reserve the term "the Institute" to refer to the organization itself.

@@

@@

The first thing to note about the IHME model is that it is atheoretical. That means that the choices that the modelers based were driven solely by what seemed to fit the data best rather than what might be expected based on how infectious diseases are transmitted.

@@

You can see some of this conflict play out in the interview that Dr. Murray has on the FiveThirtyEight website (see below). Dr. Murray argues that the theoretical models have too many assumptions and should let the actual data determine more of the model. The critics of the IHME model contend that the model changes as more data arrives, leading to unstable estimates. It's important to note here that the model itself is changing and not just the estimates produced by the model.

@@

First, the SEIR models and variations on it have been used successfully many times in the past. Every year a new flu season offers a new test of these models and every year they seem to do reasonably well.

@@

```{r g1, fig.width=4, fig.height=2}
suppressMessages(suppressWarnings(library(tidyverse)))
data.frame(
  x <- seq(-2.8, -1.5, length=100),
  y <- dnorm(x)) %>%
  ggplot(aes(x, y)) +
    geom_line() + 
    theme_void() + 
    expand_limits(x=c(-3, 0.5)) +
    expand_limits(y=c(-0.02, pnorm(0))) +
    geom_text(x=-2.5, y=-0.02, label="Time") +
    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
```

@@

Once you are past the inflection point, you can start to forecast reasonably well. You can get an initial estimate of when the number of deaths reachs its maximum and when it will die down to zero again. But these estimates are tricky and the confidence limits are very wide.

@@

@@

+ This article describes an earlier model.

@@

>>>>

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
suppressMessages(suppressWarnings(library(tidyverse)))
```

@@

To avoid too much confusion, I will use the term "IHME model" to refer to the COVID-19 prediction model developed by the Institute for Health Metrics and Evaluation and reserve the term "the Institute" to refer to the organization itself. There are a class of alternative models based on Epidemiologic principles of infectious diseases. They go by a variety of acronyms, but to simplify the language, I will refer to all of these models as SEIR models.

@@

#### The IHME model fits to deaths and not cases

The modelers made a concious decision to use the number of daily deaths to estimate the model parameters rather than the number of new daily infections. The rationale for this choice is that there are substantial variations in testing rates in different regions of the country and different countries of the world. This combined with the large number of people who are asymptomatic leads to huge uncertainties in what the number of new infections is really measuring.

The number of deaths in a region, however, is not perfect itself. Many deaths could be uncounted if they occur at home or among infected but undiagnosed patients. The flip side of the coin is that many people who were old and frail may have died with or without the infection.

There is strong empirical evidence from comparing death rates before and after the COVID-19 pandemic that we are greatly underestimating the number of deaths that we attribute to COVID-19.

Whether the underestimate of COVID-19 cases is worse than the underestimate of COVID-19 deaths is an open question. A key point to remember is that a consistent underestimate (everybody misses 25% of the cases or deaths) is not a serious problem for modeling, but a variable underestimate (some regions miss 10% and other miss 50%) is a big problem.

The alternative SEIR models generally use information from the daily infection rate to estimate parameters and buil predictions, though they can incorporate information from the

@@

The most widely criticized feature of the IHME model is that it is atheoretical. That means that the choices that the modelers based were driven solely by what seemed to fit the data best rather than what might be expected based on how infectious diseases are transmitted.

@@

You can see some of this conflict play out in the interview that Dr. Murray has on the FiveThirtyEight website (see below). Dr. Murray argues that the theoretical models have too many assumptions and that modelers should let the actual data determine more of the model. The critics of the IHME model contend that the model changes as more data arrives, leading to unstable estimates. It's important to note here that the model itself is changing and not just the estimates produced by the model.

@@

First, the SEIR models are battle-hardened. They have been used successfully many times in the past. Every year a new flu season offers a new test of these models and every year they seem to do reasonably well.

@@

```{r g1, fig.width=4, fig.height=2}
data.frame(
  x <- seq(-2.8, -1.5, length=100),
  y <- dnorm(x)) %>%
  ggplot(aes(x, y)) +
    geom_line() + 
    theme_void() + 
    expand_limits(x=c(-3, 0.5)) +
    expand_limits(y=c(-0.02, pnorm(0))) +
    geom_text(x=-2.5, y=-0.02, label="Time") +
    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
```

@@

Once you are past the inflection point, you can start to forecast further into the future. You can get an initial estimate of when the number of deaths reachs its maximum and when it will die down to zero again. But these estimates are tricky and the confidence limits are very wide.

@@

Fair enough, but the regions that are still early in the pandemic still want predictions on when they will reach their peak and when they will eventually decline to zero. The IHME model does this through a mixed model approach.

The mixed model is a well-established approach in Statistics. It is used a lot in Epidemiology, especially in small area estimation (getting estimated rates in regions that don't have a lot of data). The mixed model borrows information from other regions that are (hopefully) similar and provides much more precision than you could get by relying on the one region alone.

For the IHME model, the borrowing of information occurs mostly on the far-side (the declining side) of the daily death rate, This is especially true for those regions that still only have data on the near-side (the rising side) of the daily death rate. These regions that have data on the far-side are mostly from China, South Korea, and Italy.

Does it make sense to base the decline in the daily death rate in Chattahoochee county, Georgia on the decline in the daily death rate in Bergamo, Italy? Well, no not really, but what other choice do you have?

The theoretical models in Epidemiology provide an alternate way to extrapolate the far-side of these curves. The shape of the curves are estimated by the rates at which people move from one compartment to another in the SEIR model and variations on that model. You can get partial information on the various rates from the near-side data and then plug in information from earlier epidemics to help with predictions on the far-side.

In either the IHME model or the SEIR models require you to mix apples and oranges. In the IHME model, the oranges are places like Bergamo, Italy. In the SEIR mode, the oranges are epidemics like H1N1 influenza. The SEIR oranges, though, have one advantage of the IMHE oranges. There are many more SEIR oranges to choose from. The IMHE oranges are, at least for now, limited to China, South Korea, and Italy.

#### The IHME model has changed over time

All models provide different predictions as you accumulate more data. This leads to an inconsistent message over time and is largely unavoidable in any model. The IHME model, however, has undergone changes in the underlying mechanics that have increased its instability over time. Some of these changes were made to address criticisms in the model and some were made to get better fits to the data. One of the changes was noted above in the replacement of a single bell shaped curve with a weighted average of multiple bell shaped curves.

The desire for consistency can be overrated. Stephen Senn talks about a medical biostatistician who doesn't believe in America because it wasn't in Christopher Columbus's original research plan. Still, the changes are a legitimate concern among some of the critics of the IHME model.

### Resources

@@

+ This article describes an earlier model IHME model.

@@

Bui Q, Katz J, Parlapiano A, and Sanger-Katz, M. What 5 Coronavirus Models Say the Next Month Will Look Like. The New York Times, 2020-04-22. Available in [html format](https://www.nytimes.com/interactive/2020/04/22/upshot/coronavirus-models.html).

+ This article got me interested in the differences of the various models and has links that provide more details about the IHME model and several different SEIR models in greater detail.
